Solution of hydrodynamic equilibrium in black-oil reservoirs: the classical material balance equation

### 2014-06-30 pituba/article/article_spe.tex
improvs

An equation for the bubble-point can be obtained by setting $\bar{S}_{g2}=0$ in Eq.~\eqref{eq: Sg5} and solving it for the solubility ratio at bubble-point pressure, or $R_s(p=P_b)$:
\begin{equation}
\begin{split}
R_{s2}\left(p=P_b\right) = &\Big[N\left(m B_{1}/B_{g1} +R_{s1}\right)\\
&+G_i-G_p\Big]/\left(N-N_p\right) \, .
\end{split}
\end{equation}

@@

Similarly for the gas phase, using the state equation $pV = Zn R T$, we obtain its isothermal compressibility:
\begin{equation}
\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial p_g}\right)_{T} = \frac{1}{p_g}-\frac{1}{Z}\frac{\partial Z}{\partial p_g} \, .
\end{equation}
Neglecting the variation of the Z-factor and using hydrostatic equilibrium again, we obtain the derivative of gas density with depth as:
\begin{equation}\label{eq: gradgdz}
\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial z}\right)_{T} = \left(\frac{\mathcal{M}}{ Z R T}\right) g \, .
\end{equation}

@@

Oil and gas densities are thus approximately constant with depth.
The reference depth for computing phase density (and also the formation volume factor) can be any arbitrary depth inside phase volume. Under the approximation of a constant density, it is straightforward to show that $\bar{\rho}_\alpha = \rho_\alpha\left(\bar{p}_\alpha\right)$ and $\bar{p}_\alpha=p_\alpha\left(z^0_\alpha\right)$ where $z^0_\alpha$ is the depth of phase gravity center and $\alpha=w, o, g$. The pressure calculated in the material balance equation corresponds to the local pressure at $z^{ref}$, without fluid motion, and only equals the average reservoir pressure $\bar{p}_\alpha$ if $z^{ref}=z^0$.

For the case of an oil reservoir with a gas cap, the reference depth can be defined, with advantage, at the gas-oil contact. The point here is that at phase contact both gas and oil densities are computed using the same pressure, i.e., the corresponding local value.

If the reference depth for oil phase is otherwise defined below gas-oil contact, gas density must be computed at some different depth still inside gas phase, i.e., at or above gas-oil contact. This is to avoid the effect of oil hydrostatic gradient in the computation of gas density,
\begin{equation}
\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial z}\right)_{T,Z} = \frac{\rho_o}{\rho_g}\left(\frac{\mathcal{M}}{Z R T}\right)g  \, ,
\end{equation}
what is $\rho_o / \rho_g$ times greater than the sole effect of gas hydrostatic pressure that was computed in Eq.~\eqref{eq: gradgdz}.

It has been shown that pressure variation with depth is not important for the computation of gas, oil and water densities. Nevertheless, it was said that pressure also plays a second role in the material balance equation by defining phase equilibrium. For this purpose, pressure variation with depth may not be negligible in the evaluation of the solubility ratio: 
\begin{equation}\label{eq: dRsdz1}
\frac{1}{R_s}\frac{d R_s}{dz} = \frac{1}{R_s}\frac{d R_s}{dp_o} \frac{dp_o}{dz} = \frac{\rho_o g}{R_s}\frac{dR_s}{dp_o} \, .
\end{equation}

@@

\section{Solution of material balance equation}
A numerical solution of Eq.~\eqref{eq: MBE_short} is presented here. We shall start defining a residual function ($\varepsilon$) for which the pressure solving Eq.~\eqref{eq: MBE_short} is a root:
\begin{equation}\label{eq: residual}
\varepsilon(p_2) = \frac{E+F_i+C-F_p}{V_{p1}} \, .
\end{equation}
We proceed showing that $\varepsilon$ is a monotonic function provided some conditions on fluid properties are observed and that, as a consequence, only one root of $\varepsilon$ exists at most. Equivalently, we can say that only one solution exists for material balance equation.

@@

Residual function clearly satisfies the first two affirmatives because fluid properties are continuous everywhere and are also differentiable everywhere except for the first order phase change at bubble-point pressure.
The third affirmative is proved by rewriting $d\epsilon/dp$ as a sum of four parcels corresponding each parcel to a different phase. Each parcel is composed of two factors: current volume \textit{in-situ} and volume derivative with respect to pressure:
\begin{equation}
\begin{split}
&V_{p1}\frac{d\varepsilon}{dp}=\underbrace{\left(N-N_p\right) \left(\frac{dB_o}{dp} - B_g \frac{dR_s}{dp} \right)}_\text{oil phase}\\
&+\underbrace{\left[N R_{s1}+mN\frac{B_{o1}}{B_{g1}}+G_i-\left(N-N_p\right)R_{s2}-G_p\right] \frac{dB_g}{dp}}_\text{gas phase}\\
&+\underbrace{\left(W + W_i+W_e-W_p\right)\frac{dB_w}{dp}}_\text{water phase}\\
&\underbrace{-\frac{dV_p}{dp}}_\text{porous volume}\, .
\end{split}
\end{equation}

@@

Eq.~\eqref{eq: PVTH1} also provides a condition to check whether PVT data obtained from laboratory or empirical correlations is physically correct. It can be alternatively formulated as shown bellow in Eq.~\eqref{eq: vaporization}. It simply indicates that any vaporization must be followed by an increase in oil phase density:
\begin{equation}\label{eq: vaporization}
\frac{1}{\rho_o}\frac{d\rho_o}{dY_o} \geq Y_o \left( \frac{\rho_o}{\rho_g} - 1 \right) \, .
\end{equation}

@@

The steps to solve all material balance equation are summarized bellow:
\begin{step}
Compute bubble-point pressure for instant $t_2$.
\end{step}

Bubble-point pressure $P_{b2}$ is only a function of the values in initial instant $t_1$ and produced and injected volumes.

\begin{step}
Compute fluid pressure for instant $t_2$ finding the root of residual function ($\varepsilon$).
\end{step}

Pressure in Eq.~\eqref{eq: MBE} is a function only of fluid properties in instant $t_1$, produced and injected values and bubble-point pressure at instant $t_2$, $P_{b2}$, which was already computed in step 1.

>>>>

An equation for the bubble-point can be obtained by setting $\bar{S}_{g2}=0$ in Eq.~\eqref{eq: Sg5} and solving it for the solubility ratio at bubble-point pressure, or $R_s(p=P_b)$:
\begin{equation}\label{eq: eqPb}
\begin{split}
R_{s2}\left(p=P_b\right) = &\Big[N\left(m B_{1}/B_{g1} +R_{s1}\right)\\
&+G_i-G_p\Big]/\left(N-N_p\right) \, .
\end{split}
\end{equation}

@@

Similarly for the gas phase, using the state equation $pV = Zn R T$, we obtain its isothermal compressibility:
\begin{equation}
\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial p_g}\right)_{T} = \frac{1}{p_g}-\frac{1}{Z}\frac{\partial Z}{\partial p_g} \, .
\end{equation}
Neglecting the variation of the Z-factor and using hydrostatic equilibrium again, we obtain the derivative of gas density with depth as:
\begin{equation}\label{eq: gradgdz}
\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial z}\right)_{T} = \frac{\rho_g \vec{g}}{p_g} = \left(\frac{\mathcal{M}}{ Z R T}\right) g \, .
\end{equation}

@@

Oil and gas densities are thus approximately constant with depth.
The reference depth for computing phase density (and also the formation volume factor) can be any arbitrary depth inside phase volume. Under the approximation of a constant density, it is straightforward to show that $\bar{\rho}_\alpha = \rho_\alpha\left(\bar{p}_\alpha\right)$ and $\bar{p}_\alpha=p_\alpha\left(z^0_\alpha\right)$, where $z^0_\alpha$ is the depth of phase gravity center and $\alpha=w, o, g$. The pressure calculated in the material balance equation corresponds to the local pressure at $z^{ref}$, without fluid motion, and only equals the average reservoir pressure $\bar{p}_\alpha$ if the reference depth is the depth of the gravity center, $z^{ref}=z^0$.

For the case of an oil reservoir with a gas cap, the reference depth should be defined at the gas-oil contact. The point here is that at gas-oil contact both gas and oil densities are computed using the same pressure, its local value. If the reference depth, common to all phases, is otherwise defined below gas-oil contact, gas density will increase with depth at a higher rate because of the oil hydrostatic gradient,
\begin{equation}
\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial z}\right)_{T,Z} = \frac{\rho_o}{\rho_g}\left(\frac{\mathcal{M}}{Z R T}\right)g  \, ,
\end{equation}
what is $\rho_o / \rho_g$ times greater than the gas hydrostatic gradient used in Eq.~\eqref{eq: gradgdz}.

It has been shown that the pressure variation with depth is not important for the computation of gas, oil and water densities. Nevertheless, pressure also plays a second role in the material balance equation by defining phase equilibrium, for what the pressure variation with depth may not be negligible in the solubility ratio: 
\begin{equation}\label{eq: dRsdz1}
\frac{1}{R_s}\frac{d R_s}{dz} = \frac{1}{R_s}\frac{d R_s}{dp_o} \frac{dp_o}{dz} = \frac{\rho_o g}{R_s}\frac{dR_s}{dp_o} \, .
\end{equation}

@@

\section{Solution of material balance equation}
A numerical solution of Eq.~\eqref{eq: MBE_short} is presented here. We shall start defining a residual function ($\varepsilon$) for which the pressure solving Eq.~\eqref{eq: MBE_short} is a root:
\begin{equation}\label{eq: residual}
\varepsilon(p_2) = \frac{E+F_i+C-F_p}{V_{p1}} \, .
\end{equation}
We proceed showing that $\varepsilon$ is a monotonic function provided some conditions on fluid properties are observed and that, as a consequence, only one root of $\varepsilon$ exists at most. Equivalently, we can say that only one solution exists for the material balance equation.

@@

Residual function clearly satisfies the first two affirmatives because fluid properties are continuous everywhere and are also differentiable everywhere except for the first order phase change at bubble-point pressure.
The third affirmative is proved by rewriting $d\epsilon/dp$ as a sum of four parcels, each corresponding to a different phase (three fluid phases and the rock). For the fluid phases, each parcel is composed of two factors: current volume \textit{in-situ} and volume derivative with respect to pressure:
\begin{equation}
\begin{split}
&V_{p1}\frac{d\varepsilon}{dp}=\underbrace{\left(N-N_p\right) \left(\frac{dB_o}{dp} - B_g \frac{dR_s}{dp} \right)}_\text{oil phase}\\
&+\underbrace{\left[N R_{s1}+mN\frac{B_{o1}}{B_{g1}}+G_i-\left(N-N_p\right)R_{s2}-G_p\right] \frac{dB_g}{dp}}_\text{gas phase}\\
&+\underbrace{\left(W + W_i+W_e-W_p\right)\frac{dB_w}{dp}}_\text{water phase}\\
&\underbrace{-\frac{dV_p}{dp}}_\text{porous volume}\, .
\end{split}
\end{equation}

@@

Eq.~\eqref{eq: PVTH1} also provides a condition to check whether PVT data obtained from laboratory or empirical correlations is physically correct. It can be alternatively formulated in terms of densities as shown bellow in Eq.~\eqref{eq: vaporization}. It simply indicates that any vaporization must be followed by an increase in oil phase density:
\begin{equation}\label{eq: vaporization}
\frac{1}{\rho_o}\frac{d\rho_o}{dY_o} \geq Y_o \left( \frac{\rho_o}{\rho_g} - 1 \right) \, .
\end{equation}

@@

The steps to solve the material balance equation are summarized bellow:
\begin{step}
Compute bubble-point pressure for instant $t_2$ using Eq.~\eqref{eq: eqPb}.
\end{step}

Bubble-point pressure $P_{b2}$ is only a function of values at instant $t_1$ and produced and injected volumes.

\begin{step}
Compute pressure for instant $t_2$ by finding the root of residual function $\varepsilon \left(p\right)$.
\end{step}

Pressure in Eq.~\eqref{eq: MBE} is only a function of fluid properties in instant $t_1$, produced and injected values and bubble-point pressure at instant $t_2$, $P_{b2}$, which was already computed in step 1.
